### 2019-04-24 sections/abstract.tex
rename main tex file to main
rename calculus to language

% Approach
In this paper we give a description of the TOP paradigm and then decompose its rich features into elementary language elements, which makes them suitable for formal treatment.
We use the simply typed lambda calculus, extended with pairs and references, as a base language.
On top of this language, we develop TopHat, a calculus for modular interactive workflows.
We describe TopHat by means of a layered semantics.
These layers consist of multiple big-step evaluations on expressions, and two labelled transition systems, handling user inputs.

>>>>

% Approach
In this paper we give a description of the TOP paradigm and then decompose its rich features into elementary language elements, which makes them suitable for formal treatment.
We use the simply typed $lambda$-calculus, extended with pairs and references, as a base language.
On top of this language, we develop TopHat, a language for modular interactive workflows.
We describe TopHat by means of a layered semantics.
These layers consist of multiple big-step evaluations on expressions, and two labelled transition systems, handling user inputs.

### 2019-04-24 sections/conclusions.tex
Updated reference to TFP paper naus/jeuring
convert back to two column layout
rename main tex file to main
rename main file to main
rename calculus to language

In this paper we have identified and intuitively described the essence of task-oriented programming.
We then formalized this essence by developing a domain-specific calculus for declarative interactive workflows, called \TOPHAT.
The task calculus and the host language are clearly separated, to make explicit where the boundaries are.
The semantics of the task layer is driven by user input.
We have compared \TOPHAT with workflow moddeling languages, process algebras, functional reactive programming and session types to point out differences and similarities.
Finally, we have proven type safety and progress for our language.

@@

One of the main motivations to formalize task-oriented programming is to be able to reason about programs.
In this paper we reason about the language itself, but it would be nice to prove properties about individual programs.
To this end, we are very interested to see if it is possible to develop an axiomatic semantics for \TOPHAT that allows us to do so.
There are certain properties of our calculus that make this particularly complex:
We have to deal with parallelism, user interaction, and references.

@@

Another form of reasoning about programs is static analysis.
\citet{conf/ifl/KlinikJP17} have developed a cost analysis for tasks that require resources in order to be executed.
This analysis was developed for a simpler task calculus, and could be brought over to the one developed here.

\citet{UUCS2017013} have looked at building a generic feedback system for rule-based problems.
A workflow system typically is rule based, as outlined in their work.
It would be interesting to fit the generic feedback system to \TOPHAT in order to support end-users working in applications developed in this calculus.

>>>>

In this paper we have identified and intuitively described the essence of task-oriented programming.
We then formalized this essence by developing a domain-specific language for declarative interactive workflows, called \TOPHAT.
The task language and the host language are clearly separated, to make explicit where the boundaries are.
The semantics of the task layer is driven by user input.
We have compared \TOPHAT with workflow moddeling languages, process algebras, functional reactive programming and session types to point out differences and similarities.
Finally, we have proven type safety and progress for our language.

@@

One of the main motivations to formalize task-oriented programming is to be able to reason about programs.
In this paper we reason about the language itself, but it would be nice to prove properties about individual programs.
To this end, we are very interested to see if it is possible to develop an axiomatic semantics for \TOPHAT that allows us to do so.
There are certain properties of our language that make this particularly complex:
We have to deal with parallelism, user interaction, and references.

@@

Another form of reasoning about programs is static analysis.
\citet{conf/ifl/KlinikJP17} have developed a cost analysis for tasks that require resources in order to be executed.
This analysis was developed for a simpler task language, and could be brought over to the one developed here.

\citet{DBLP:conf/sfp/NausJ16} have looked at building a generic feedback system for rule-based problems.
A workflow system typically is rule based, as outlined in their work.
It would be interesting to fit the generic feedback system to \TOPHAT in order to support end-users working in applications developed in this language.

### 2019-04-24 sections/example.tex
rename main tex file to main
shorten code examples
introduce fail symbol in example
rename calculus to language

Now we develop our workflow in a top-down manner.
Our flight booking starts with an interactive task \TS{enter (List Passenger)}, where users can enter a list of passengers.
A task $\Enter \tau$ is an empty editor that asks for a value of the given type $\tau$.
Passengers are valid if their name is not empty and their age is at least 0.
Lists of passengers are valid if each passenger is valid, and at least one of the passengers is an adult.
When the user has entered a valid list of passengers, the step after \TS{>>?} becomes enabled,
and the user can proceed to picking seats.
\begin{TASK}
  let valid = \p. not (fst p == "") /\ snd p >= 0 in
  let adult = \p. snd p >= 18 in
  let allValid = \ps. all valid ps /\ any adult ps in
  let bookFlight = enter (List Passenger) >>? \ps.
    if allValid ps then chooseSeats ps else fail
\end{TASK}
A selection of seats is correct if every entered seat is free.
\begin{TASK}
  let correct = \ss. intersect ss !freeSeats == ss in
  let chooseSeats = \ps. enter (List Seat) >>? \ss.
    if correct ss /\ length ps == length ss then confirmBooking ps ss else fail
\end{TASK}

@@

All instances of the \TS{bookFlight} task have access to the shared list of free seats.
Rewriting the example in a calculus without side effects would not only be cumbersome,
obfuscating the code with explicit threading of state,
but it would be impossible to model the parallel execution of three \TS{bookFlight} tasks.
It is not known upfront which task will finish first,
and thus it is not possible to thread the free seat list between the parallel tasks.

>>>>

Now we develop our workflow in a top-down manner.
Our flight booking starts with an interactive task \TS{enter (List Passenger)}, where users can enter a list of passengers.
A task $\Enter \tau$ is an empty editor that asks for a value of the given type $\tau$.
Passengers are valid if their name is not empty and their age is at least 0.
Lists of passengers are valid if each passenger is valid, and at least one of the passengers is an adult.
When the user has entered a valid list of passengers, the step after \TS{>>?} becomes enabled,
and the user can proceed to picking seats.
In case of an invalid list of passengers,
the step is guarded by the failing task \TS{fail}.
\begin{TASK}
  let valid = \p. not (fst p == "") /\ snd p >= 0 in
  let adult = \p. snd p >= 18 in
  let allValid = \ps. all valid ps /\ any adult ps in
  let bookFlight = enter (List Passenger) >>? \ps.
    if allValid ps then chooseSeats ps else fail
\end{TASK}
A selection of seats is correct if every entered seat is free.
\begin{TASK}
  let correct = \ss. intersect ss !freeSeats == ss in
  let chooseSeats = \ps. enter (List Seat) >>? \ss.
    if correct ss /\ length ps == length ss then confirmBooking ps ss else fail
\end{TASK}

@@

All instances of the \TS{bookFlight} task have access to the shared list of free seats.
Rewriting the example in a language without side effects would not only be cumbersome,
obfuscating the code with explicit threading of state,
but it would be impossible to model the parallel execution of three \TS{bookFlight} tasks.
It is not known upfront which task will finish first,
and thus it is not possible to thread the free seat list between the parallel tasks.

### 2019-04-24 sections/introduction.tex
rename main file to main
rename calculus to language

In this paper, we want to take a step back and look at the spirit of \TOP.
We do this both formally and informally.
Informally in the sense that we give an intuitive description of the features that define task-oriented programming.
Formally in the sense that we develop a calculus which formalises these features as language constructs,
and we give them a semantics in the style that is common in programming language research.
We separate the task layer and the underlying host language, both syntactically and semantically.
Thus making explicit which properties of \TOP come from the task layer, and which come from functional programming.
Our challenge, therefore, is to model the properties of \TOP into a calculus
and pave the way for formal treatment of \TOP programs.
We give this formal calculus the name \TOPHAT (TopHat).

@@

  \item
    We present a formal calculus for modelling declarative workflows, embedded in a simply typed lambda-calculus, called \TOPHAT.
    It is based on the aforementioned essential \TOP concepts.

@@

  \item
    Using both the essential concepts and the formal calculus, we compare \TOP with related work in areas ranging from business process modelling, to process algebras and reactive programming.

@@

In \cref{sec:example} we demonstrate the functionality of \TOPHAT by means of an example,
\Cref{sec:intuition} gives an overview of the essential concepts of \TOP.
\Cref{sec:language} introduces the \TOPHAT calculus syntax
and \cref{sec:semantics} the semantics.
Then in \cref{sec:properties} we show that certain properties hold for the calculus.
We take a look at related work in \cref{sec:relatedwork}
and conclude in \cref{sec:conclusions}.

>>>>

In this paper, we want to take a step back and look at the spirit of \TOP.
We do this both formally and informally.
Informally in the sense that we give an intuitive description of the features that define task-oriented programming.
Formally in the sense that we develop a language which formalises these features as language constructs,
and we give them a semantics in the style that is common in programming language research.
We separate the task layer and the underlying host language, both syntactically and semantically.
Thus making explicit which properties of \TOP come from the task layer, and which come from functional programming.
Our challenge, therefore, is to model the properties of \TOP into a language
and pave the way for formal treatment of \TOP programs.
We give this formal language the name \TOPHAT (TopHat).

@@

  \item
    We present a formal language for modelling declarative workflows, embedded in a simply typed $lambda$-calculus, called \TOPHAT.
    It is based on the aforementioned essential \TOP concepts.

@@

  \item
    Using both the essential concepts and the formal language, we compare \TOP with related work in areas ranging from business process modelling, to process algebras and reactive programming.

@@

In \cref{sec:example} we demonstrate the functionality of \TOPHAT by means of an example,
\Cref{sec:intuition} gives an overview of the essential concepts of \TOP.
\Cref{sec:language} introduces the \TOPHAT language syntax
and \cref{sec:semantics} the semantics.
Then in \cref{sec:properties} we show that certain properties hold for the language.
We take a look at related work in \cref{sec:relatedwork}
and conclude in \cref{sec:conclusions}.

### 2019-04-24 sections/language.tex
make grammars fit columns
make rules fit columns
exclude editor figure
rename main file to main
rename calculus to language

In this section, we present the constructs of \TOPHAT, our modular interactive workflow calculus.
We define the host and task language, the types, and the static semantics.
Then we describe the workings of each construct using examples.
These constructs are formalised in \cref{sec:semantics}.

@@

\label{sub:expressions}
The host language is a simply typed lambda calculus, extended with some basic types and \ML-style references.
We use references to represent shared data sources.
The grammar in \cref{fig:language-grammar} defines the syntax of the host language.
It has abstractions, applications, variables, and constants for booleans, integers and strings.
The symbol $\star$ stands for binary operators.
For the result of parallel tasks we need pairs.
Conditionals come in handy for defining guards.
%
References will be used to implement shared editors.
Our treatment of references closely follows the one by \citet{books/Pierce02TAPL}.
Creating a reference using the keyword $\Ref$ yields a location $l$.
While $x$ denotes program variables, $l$ denotes store locations.
Locations are not intended to be directly manipulated by the programmer.
The symbols ! and $:=$ stand for dereferencing and assignment.
The unit value will be used as the result of assignments.

@@

% \subsection{Appointment}
%
%
% It is essential for a workflow system to support multiple collaborators working
% together. Adding this funcitonaltiy to the \TOPHAT calculus is very
% straightforward. We identify users by means of a name $u$ which we can then use
% to label appointed tasks. The sole purpose of this labelling is to be able to
% tell which inputs are expected from what users. Other than that, it has no
% influence on the semantics. For space reasons, we have have excluded multi-user
% support from this paper.
%
% Mss moeten we het schrijven vanuit het oogpunt van labeling: je kunt labels
% toevoegen aan knoppen, als uitleg bij taken en om gebruikers toe te kennen aan
% taken. Vervolgens zou je iets met die gebruikers kunnen doen: alleen taken voor
% hun weergeven in de UI, alleen inputs voor hun genereren met I etc. Past dat?

>>>>

In this section, we present the constructs of \TOPHAT, our modular interactive workflow language.
We define the host and task language, the types, and the static semantics.
Then we describe the workings of each construct using examples.
These constructs are formalised in \cref{sec:semantics}.

@@

\label{sub:expressions}
The host language is a simply typed $lambda$-calculus, extended with some basic types and \ML-style references.
We use references to represent shared data sources.
The grammar in \cref{fig:language-grammar} defines the syntax of the host language.
It has abstractions, applications, variables, and constants for booleans, integers and strings.
The symbol $\star$ stands for binary operators.
For the result of parallel tasks we need pairs.
Conditionals come in handy for defining guards.
%
References will be used to implement shared editors.
Our treatment of references closely follows the one by \citet{books/Pierce02TAPL}.
Creating a reference using the keyword $\Ref$ yields a location $l$.
While $x$ denotes program variables, $l$ denotes store locations.
Locations are not intended to be directly manipulated by the programmer.
The symbols ! and $:=$ stand for dereferencing and assignment.
The unit value will be used as the result of assignments.

@@

% \subsection{Appointment}
%
%
% It is essential for a workflow system to support multiple collaborators working
% together. Adding this funcitonaltiy to the \TOPHAT language is very
% straightforward. We identify users by means of a name $u$ which we can then use
% to label appointed tasks. The sole purpose of this labelling is to be able to
% tell which inputs are expected from what users. Other than that, it has no
% influence on the semantics. For space reasons, we have have excluded multi-user
% support from this paper.
%
% Mss moeten we het schrijven vanuit het oogpunt van labeling: je kunt labels
% toevoegen aan knoppen, als uitleg bij taken en om gebruikers toe te kennen aan
% taken. Vervolgens zou je iets met die gebruikers kunnen doen: alleen taken voor
% hun weergeven in de UI, alleen inputs voor hun genereren met I etc. Past dat?

### 2019-04-24 sections/relatedwork.tex
rename main tex file to main
rename main file to main
rename calculus to language

\BPEL~\cite{bpel} is another popular business process language. The standardised
language allows for the specification of actions within business processes,
using an \XML format.
The language is mainly used for coordinating web services.
Two workflow patterns are supported; execution of services can be done sequential or in parallel.
On top of that, processes can be guarded by conditionals.
There is no support for higher order processes however.
Processes described in \BPEL can be regarded as activity graphs, and they can also be rendered as such.
The specified processes in \BPEL are directly executable, just like \YAWL.

@@

As with \TOPHAT, HipHop is a \DSL embedded in a general purpose programming language.
% the same way \TOPHAT is embedded into the simply typed $\lambda$-calculus.
Another similarity is that both specifications lead to executable server and client applications from a single source.
However, both HipHop and Esterel are more low level regarding their specification.
Where \TOPHAT takes tasks and collaboration as a starting point,
HipHop focusses on synchronous communication and atomic execution of reactive machines.

@@

Chorus~\cite{chen2017chorus} is a visual programming environment for online
collaboration apps. The goal of Chorus is to allow users to program their own
application, though which a collective task can be coordinated. Users work on a
preset data-type that drives the behaviour of the user defined applications.
\fixme{compare with tophat}

>>>>

\BPEL~\cite{bpel} is another popular business process calculus. The standardised
language allows for the specification of actions within business processes,
using an \XML format.
The language is mainly used for coordinating web services.
Two workflow patterns are supported; execution of services can be done sequential or in parallel.
On top of that, processes can be guarded by conditionals.
There is no support for higher order processes however.
Processes described in \BPEL can be regarded as activity graphs, and they can also be rendered as such.
The specified processes in \BPEL are directly executable, just like \YAWL.

@@

As with \TOPHAT, HipHop is a \DSL embedded in a general purpose programming language.
% the same way \TOPHAT is embedded into the simply typed $\lambda$-language.
Another similarity is that both specifications lead to executable server and client applications from a single source.
However, both HipHop and Esterel are more low level regarding their specification.
Where \TOPHAT takes tasks and collaboration as a starting point,
HipHop focusses on synchronous communication and atomic execution of reactive machines.

@@

Chorus~\cite{chen2017chorus} is a visual programming environment for online
collaboration apps. The goal of Chorus is to allow users to program their own
application, though which a collective task can be coordinated. Users work on a
preset data-type that drives the behaviour of the user defined applications.
Chorus can not be regarded as a fully fledged workflow system, nor does it aim
to be. The data-type that drives the behaviour of the app does not change during
execution, and therefore heavily limits what flows can be encoded.
